From Active Hostilities to Frozen Stalemates: Assessing the Role of External Support in Civil Conflict Dynamics

### 2023-11-08 doc/coding-protocol.qmd
Move dataset docs
Manuscript notes

@@

```{r echo = F}
candidates <- read.csv("./data/dyadic_candidates.csv")
episodes <- read.csv("./data/dyadic_episodes.csv")
```

>>>>

TODO: manual coding, specific criteria for determining when an actor ceases to exist
TODO: how to handle actor fragmentation? Ex: Sendero Luminoso
TODO: Be specific about sources that will be used, reference list?

@@

```{r echo = F}
candidates <- read.csv("../data/dyadic_candidates.csv")
episodes <- read.csv("../data/dyadic_episodes.csv")
```

### 2023-11-08 paper.qmd
Manuscript notes

[//]: # Setup R environment
```{r setup, include = F}
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggdist)
library(modelsummary)
library(knitr)
library(patchwork)
library(tidyr)

@@

@@

In what follows, I will first describe the dependent and independent variables, including operationalizations and data sources. I will then present the model specification and results from a conventional probit regression followed by the instrumental variable analysis.

@@

### Defining Frozen Conflicts

@@

UCDP external support dataset

@@

- Conflict Intensity
- Conflict Duration
- Incompatibility?
- e_gdppc
- e_pop
- CINC

@@

@@

@@

## Robust Checks

### Observational Probit Models

>>>>

[//]: # Setup R environment
```{r setup, include = F}
library(dplyr)
library(flextable)
library(ggplot2)
library(ggthemes)
library(ggdist)
library(modelsummary)
library(knitr)
library(patchwork)
library(tidyr)

@@

**Motivation**

Contribution: (*conceptual* and *empirical*)

- *Original Data*, conceptual/descriptive claim about prevalence of frozen conflicts
- Novel empirical analysis, provide a more nuanced analysis of *conflict termination* and contributes to our understanding of how external support can shape the dynamics of conflicts

The paper will argue that there exists both a theoretical and empirical gap. Studies on conflict termination have generally considered military victory and peace agreements as the only outcome. I further disaggregate the notion of termination to cover also the situation where conflicts are not conclusively termination and instead become "frozen."

In doing so, I show that frozen conflicts are more prevalent than thought and conceptually can and should cover a greater range of cases than just /de facto/ states. I also distinguish the concept of frozen conflicts from the related notions of stalemates and protracted conflicts (and maybe intractable rivalries?).

I provide a novel quantitative analysis on the likelihood of a conflict becoming frozen. Specifically, I examine the effect of support by external state actors to belligerents in the course of a conflict episode. This will contribute both to our understanding of different types of conflict termination as well as the ways in which external support can shape the dynamics of conflicts.

- Potential shortcoming: what I am not doing is taking into account the preferences of external supporters and motivation for offering support. Nor am I distinguishing between different external support types.

**Why should we care about frozen conflicts?**

# Defining Frozen Conflicts

# Theory

*The fundamental question*: why would actors accept a frozen conflict over a comprehensive peace agreement?

This mirrors the "why fight" question underlying bargaining theory where fighting is assumed to be costly such that with perfect information rational actors should prefer a "bargained solution." Is there a similar parallel to frozen conflicts?

What is the cost of perpetuating a frozen conflict? How does this play in with the traditional, bargaining theory approach to conflict termination. Namely, actors become willing to terminate a conflict as information is revealed through fighting and the payoff from victory diminishes, based on the subjective probability of winning and the accumulated costs both past and future.

Is there a situation in which rational actors would prefer to leave a conflict unresolved with the future threat of violent re-escalation? Clearly, the rebel side faces a credible commitment dilemma in that they can never guarantee that the government won't renege in the future.

Literature on ceasefires (ex Clayton and Stitcher (2021)) emphasize the strategic logic underlying actors' decisions to accept or reject a ceasefire. More specifically, actors may be willing to enter into a ceasefire either as a signaling mechanism to demonstrate an intent to peacefully resolve the conflict or as a delaying mechanism to allow them to temporarily pause fighting in order to rebuild their strength. The decision to accept a frozen conflict, despite the ongoing costs, may be similar to this latter mechanism. Conflict actors may not be willing to completely abandon their war aims; however, due to a variety of factors are unable to win an outright victory. By letting the conflict become frozen, either through a ceasefire, partial peace agreement, or a stalemate, actors can temporarily halt the ongoing costs of fighting in hopes of a future favourable shift in the balance of power. This probably requires actors to have long time horizons and be willing to burden the costs of not fully resolving the conflict. Specifically, letting a conflict become frozen means that there is always a threat of re-escalation so actors need to continue to commit to military spending, while suffer ongoing economic costs from lost territory.

"...Leeds (1999) and Walter (2002) suggest that once negotiation begins because of a hurting stalemate, any warring group would prefer to have those negotiations be successful rather than to return to their group members without a settlement." --- from Park (2015)

This is a potential theoretical contribution to the bargaining literature in that I could argue a comprehensive peace agreement does not necessarily always result from a hurting stalemate. Letting conflicts freeze could be one way that actors ease the costs of a hurting stalemate without being forced to commit to a settlement. This could be especially attractive in scenarios where there is such a high degree of mutual distrust, with few international guarantees for a negotiated settlement, where credible commitment problems could lead to a frozen conflict. Especially when actors receive external support, which a) is more likely to lead to the type of stalemate scenario where neither side can outright win militarily and b) allow actors to sustain the military threat required for a frozen conflict to perpetuate.

Note, extensive literature arguing that external actors also increase the number of veto players which decrease the likelihood of a negotiated settlement (ex Cunningham 2006).

- Differentiating the theoretical mechanisms for the different treatment conditions: no external support, both sides receive external support, government only, rebel only

- States tend to be fairly conservative about sovereignty claims (see Hutchinson f.ex.). The best that separatists can hope for is a *de facto* state.

- Big difference compared to ceasefires, frozen conflicts are not /a priori/ agreed upon by belligerents. Even if an actor prefers a conflict to become frozen, there is no guarantee that the opposing side won't attempt a re-escalation at any time.

  - Even if actors prefer a frozen conflict over a comprehensive peace agreement; why not codify the status quo and formalized a ceasefire? (I think the key here is to emphasize that actors **cannot** win militarily, in other words they are stalemated into a frozen conflicts and refuse to give up their war aims. The question then is not, why do conflict actors prefer frozen conflicts, but rather in the face of a frozen conflict, why do conflict actors refuse a negotiated settlement?)

@@

In what follows, I will first present the model specification and results from a probit regression predicting the onset of a frozen conflict. I will then decrease the dependent and independent variables, including operationalizations and data sources.

## Probit Model

TODO: eventually I would like to move towards a dyadic analysis, rather than conflict level

- Unit of analysis: conflict episodes, ie consecutive years of active fighting (above UCDP threshold of 25 battle deaths)
- Binary outcome, binary treatment, while conditioning on a number of relevant covariates

- Estimate with Stan, weakly informative priors, $\delta \sim \text{Normal}(1, 1)$

@@

@@

UCDP external support dataset, binary variable for whether an external
state actor offered either fungible or non-fungible support at least
once to either the rebels or government side in a conflict episode.

- Include interaction term, three treatment conditions:

1. No support (control group)
2. Gov. support only
3. Reb. support only
4. Both support

@@

- Episode Intensity
- Episode Duration
- Incompatibility? (ie separatist conflict over territory)
- Recurrent episode
- Right censored

@@

Figure 1 shows the parameter estimates for $\delta$.

@@

Figure 2 shows the average marginal effects, *i.e.* the change in the posterior predicted probability between each treatment condition and the control condition averaged across the sample.

\begin{equation}
\text{AME}_j = \sum_{i=1}^N \Phi(\alpha + X_i' \beta + \delta_i) - \Phi(\alpha + X_i' \beta)
\end{equation}

@@

## Robustness Checks

Different combinations of the following...

- Aggregation of the treatments: binary treatment, binary treatment in the last 5 years of the conflict episode, and proportional treatment. The latter is calculated as the proportion of conflict-years in an episode where one side received external support

- All conflict episodes and only high intensity episodes --- try to distinguish between low-level insurgencies and actual civil wars

- Normal definition for frozen conflicts and a strict definition that requires a duration of at least 2 years in order to exclude temporary one-year pauses in fighting that may be more the result of UCDP's coding decision rather than representing actual frozen conflict.
